---
title: "Exercise 6.1: Linear regression on the fish tank dataset - solution"   
author: "Lieven Clement and Jeroen Gilis"
date: "statOmics, Ghent University (https://statomics.github.io)"  
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

# Fish tank dataset

In this experiment, 96 fish (dojofish, goldfish and zebrafish)
were placed separately in a tank with two liters of water and
a certain dose (in mg) of a the poison EI-43,064. The resistance
of the fish against the poison was measured as the amount of
minutes the fish survived after being exposed to the poison (`Surv_time`, in
minutes). Additionally, the weight of each fish was measured.

# Goal

The research goal is to study the association between the dose of
the poison that was administered to the fish and their
survival time.

Read the required libraries

```{r, message = FALSE}
library(tidyverse)
```

# Import the data

```{r, message=FALSE}
poison <- read_csv("https://raw.githubusercontent.com/statOmics/PSLS21/data/poison.csv")
```

# Data tidying

We can see a couple of things in the data that can be improved upon:

1. Capitalise the fist column name. *Hint*: use the `rename` function.

2. Set the Species column as a factor. *Hint*: use the `mutate` function.

3. Change the species factor levels from 0, 1 and 2 to
Dojofish, Goldfish and Zebrafish. *Hint*: use the `fct_recode` function inside
the `mutate` function.

```{r}
poison <- poison %>%
  rename("Species" = "species") %>%
  mutate(Species = as.factor(Species)) %>%
  mutate(Species = fct_recode(Species, Dojofish = "0", Goldfish = "1", Zebrafish = "2"))

poison
```

# Data exploration

How many fish do we have per species?

```{r}
poison %>%
  count(Species)
```

Make a suitable visualization of the association between
the dose and the survival time. Additionally, add the fish species
as a color to the plot.

```{r}
poison %>%
  ggplot(aes(x=Dose,y=Surv_time)) +
  geom_point() + 
  stat_smooth(method = "loess") +
  geom_smooth(method='lm',col="black") + 
  ylab("Survival time (min)") +
  xlab("Dose (mg)") +
  geom_point(aes(col=Species)) + ## not necessary, additional layer of information
  scale_color_manual(values = c("red","darkgoldenrod","black")) +
  theme_bw()
```

The linear regression line (black) is a good approximation
of the best fitting smooth line (blue) through the data.
Based on this figure, it seems realistic to suggest a linear
relationship between dose and survival, where higher doses
have lower survival times (as expected). However, before we
start with the regression analysis we must check if all required
assumptions are met.

# Important note on the dataset

In this dataset, there are multiple variables can have an effect on the survival
time of the fish. The most obvious one is the dose of poison that was 
administered (as displayed above). However, we could also imagine that heavier 
fish are less prone to the poison than light fish. Additionally, one fish
species may be more resistant to the poison than the other.

For instance, not taking into account species into the analysis will introduce 
problems regarding the assumption of *independence* of the data. The assumption 
of *independence* entails that know the response value of one observation (fish) 
does not learn us anything about the response values for another variable.
This will not be the case here; the response values (survival times) of fish
of the same species will be much more alike compared to those of fish from
another species.

Thus, in order to correctly analyze this data, fish `species` and `weight` 
should be taken into consideration. **For now, we will "avoid" taking into**
**account the effect of species into the analysis by only performing an**
**analysis for the dojofish**. As such, we will filter the dataset so that
it only contains observations for the Dojofish. 

```{r}
# filter the data so it only contains dojofish
poison_dojo <- poison %>%
    filter(Species == "Dojofish")
```

Later in the course, we will come back to this dataset and perform a fully 
correct analysis that allows us to assess the different research hypotheses for 
all fish species at once!

The second predictor variable, `Weight`, still may influence our analysis; fish
with higher weight are expected to be more resistant to the poison. **If** there
would be a systematic bias in the experiment, e.g. if all lighter fish obtained
a low dose of poison of poison and heavier fish obtained a high dose of poison
(so, if the dose was not correctly randomized accros the different weights),
then we would not be able to correctly estimate the linear association between
`Dose` and `survival`; indeed, this relationship would be confounded by the 
effect of weight.

```{r, message=FALSE}
poison_dojo %>%
  ggplot(aes(x = Dose, y = Weight)) +
  geom_point() +
  ggtitle("Association between dose and weight") +
  theme_bw() + 
  stat_summary(
    geom = "point",
    fun.y = "mean",
    col = "black",
    size = 4,
    shape = 24,
    fill = "red")
```

In our dataset, there seems to be **no systematic bias**, i.e., no clear trend 
in the association between `Dose` and `Weight`. However, we do notice that the 
fish weights are not perfectly uniformly distributed across the different doses;
we don't see a clear trend, but we do see some fluctuations. To check if 
fluctuations of the size that we observe here could happen when in reality there
is no association between `Dose` and `Weight`, we may use simulations:

```{r}
library(gridExtra)
```

```{r}
set.seed(1351)

plots <- lapply(1:9,
       function(x) poison_dojo %>%
         mutate(Dose=sample(Dose))%>% # randomization
         ggplot(aes(x = Dose, y = Weight)) +
         geom_point() +
         theme_bw()  + 
         stat_summary(
            geom = "point",
            fun.y = "mean",
            col = "black",
            size = 2,
            shape = 24,
            fill = "red")
          )
do.call(gridExtra::grid.arrange, plots)
```

In the above simulation, we randomly assigned the different fish to different 
poison doses. Even though we did this randomly, we still see that by random 
chance there can be fluctuations as large as the ones observed in our real 
dataset. As such, we may conclude that there is indeed **no systematic bias**, 
i.e., no clear trend in the association between `Dose` and `Weight`.

# Linear regression analysis

## Assumptions of linear regression

List assumptions:

1. The observations are independent of each other
2. Linearity between the response and predictor variable
3. The residues of the model must be normally distributed
4. Homoscedasticity of the data

The first assumption is **met** based on the experimental design. In the
data exploration phase, we saw that there is no clear trend in the association 
between `Dose` and `Weight`. With respect to species, we will "avoided" taking 
into account the effect of species into the analysis by only performing an
analysis for the dojofish.

We will check the other assumptions by first fitting the linear
model and plotting the output. As such, we will get all the
required diagnostic plots.

```{r}
# fit a linear regression model with 'Surv_time' as response variable and
# 'Dose' as predictor variabele
model <- lm(Surv_time~Dose, data=poison_dojo) 

## display the diagnostic plots of the model
plot(model)
```

We have four diagnostic plots:

2. Linearity with the Residuals vs fitted plot

- predictor of predictions $\hat\beta_0+\hat\beta_1 x$ on $X$-axis
- *residuals* on $Y$-as
$$e_i=y_i-\hat{g}(x_i)=y_i-\hat\beta_0-\hat\beta_1\times x_i,$$

If there would be a linear relationship in the data,
the residuals are expected to lie on the y=0 line for
the entire range of predicted values. This is clearly
the case: the assumption of linearity is met.

3. Normal Q-Q

- QQ-plot of the residuals $e_i$.

The residuals of the linear regression model should be normally
distributed. Given the second diagnostic plot, this seems not
to be the case. In stead, we observe a short left tail and a
long right tail. The assumption of normal residues is **not met.**

4. Homoscedasticity

- Square-root of the absolute value of standardized residuals
in function of the fitted values

To meet the third assumption of linear regression, the variance
on the _Square-root of the absolute value of standardized residuals_
must be similar over the entire range of fitted values. The smoother
in the plot helps us with looking at this; it should be nicely
horizontal over the entire range of fitted values. This seems not
to be the case: for larger fitted values, the variance increase slightly.
The assumption of homoscedasticity is **not met.**

Multiple assumptions are not met, so way not continue by performing
a linear regression.

One "trick" is to log-transform the data. Indeed, data with heavy
right tails can often be normalized by applying this strategy

Generate a new column in the poison dataset, containing the 
log-transformed survival times.

```{r}
poison_dojo <- poison_dojo %>%
  mutate(log.Surv_time = log(Surv_time))
```

Refit the (log-)linear model. Generate diagnostic plots.
Assess the assumptions.

```{r}
log.model <- lm(log.Surv_time~Dose, data=poison_dojo) 
plot(log.model)
```

Arguably, there might be some deviation from normality in the left tail of the
distribution. However, when we would simulate data under the normality
assumption, it seems that deviations of this size may be expected when
normality is met:

```{r}
set.seed(1406)
nobs <- nrow(poison_dojo)

data.frame(
  y = c(log.model$res,
        rnorm(nobs*8,
              sd = sigma(log.model)
             )
      ),
  label = rep(
              c("orig",
                paste0("sim",1:8)),
              each = nobs)) %>%
  ggplot(aes(sample = y)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~ label)
```

1. The independence assumption is not met, but for this exercise,
we will assume it **is met** (see above).
2. Upon log-transformation, the linearity assumption is still **met**.
3. Upon log-transformation, the normality assumption is also **met**.
4. Upon log-transformation, the homoscedasticity assumption is also **met**.

Look at the output of the log-linear model:

```{r}
summary(log.model)
```

Compute the 95% confidence interval on the model parameters:

```{r}
confint(log.model)
```

## Interpretation on the log scale 

Currently, all the outcomes should be interpreted on the log-scale.
Indeed, since we are now modelling  _the log of survival time_, we 
don't have direct inference on survival time.

As such, we may interpret the output as follows:

- The slope:

If the dose at which the poison is administered increases with 1 mg,
then the mean of the natural logarithm of survival time will decrease
by 0.5112 units (95% CI [-0.8109,-0.2116]). This decrease is significant
on the 5% significance level (p=0.00103).

- The intercept:

The average of the natural logarithm of survival time for fish that
were given a dose of 0 mg is 2.1046 (95% CI [1.6288, 2.5804]). This
value is significantly different from 0 on the 5% significance level
(7.13e-14). This conclusion will be discussed in more detail below.

## Interpretation on the original scale 

The interpretation the log-scale is quite difficult: the natural 
logarithm survival time is not exactly a casual measure. To ease 
the interpetation, we may backtransform the results to the original
scale (time in minutes). This we can do by taking the exponent of
the outcomes:

```{r}
exp(summary(log.model)$coefficients[,"Estimate"])
```

and of their confidence intervals:

```{r}
exp(confint(log.model))
```

Now, we can interpret the results in terms of the geometric mean:

- The slope:

If the dose at which the poison is administered increases with 1 mg,
then the geometric mean of survival time will decrease by a factor
of 0.5998 (95% CI [0.4445,0.8093]). This decrease is significant
on the 5% significance level (p=0.00103).

- The intercept:

The geometric mean of the survival time for fish that were given a 
dose of 0 mg is 8.2039 minutes (95% CI [5.0977, 13.2029]). This
value is significantly different from 0 on the 5% significance level
(7.13e-14).

To go a little further on the latter interpretation, our analysis seems
to suggest that fish in uncontaminated water would only suvive for about
8 minutes! This seems a little bit _fishy_...

Here, we make a diagnostic plot to explain this behavior:

```{r}
plot(x=poison_dojo$Dose, 
     y=poison_dojo$Surv_time, 
     xlab="Dose (mg)", 
     ylab="Survival time (min)", 
     xlim=c(0,2.2))
abline(lsfit(poison_dojo$Dose,poison_dojo$Surv_time), lty=2) ## lty=2 draws a dashed line

## add the intercept as a point
summary_model <- summary(model)
points(0,summary_model$coefficients["(Intercept)","Estimate"],col="red", pch=16)
```

We can see that the dose value of 0 mg does not fall within
the same range as our data. As such, it is unlikely that we
can assume this model to have correct inference on the intercept:
we expect that the linearity assumption will not be met for these
low values. The model will only be trustworthy on the range of
data that we use to build the model on!

This is not really a problem: we are mainly interested in the slope
parameter in the given range of datapoints.

# Conclusion

- On the log scale:

If the dose at which the poison is administered increases with 1 mg,
then the mean of the natural logarithm of survival time will decrease
by 0.5112 units (95% CI [-0.8109,-0.2116]). This decrease is highly
significant on the 5% significance level (p=0.00103).

- On the original scale:

If the dose at which the poison is administered increases with 1 mg,
then the geometric mean of survival time will decrease by a factor
of 0.5998 (95% CI [0.4445,0.8093]). This decrease is highly significant
on the 5% significance level (p=0.00103).

n the tutorial of multiple regression, we will revisit this exercise. 
More specifically, we will there study the association between dose and survival
for all the different species.










