---
title: "Exercise 7.2: ANOVA on the cuckoo dataset - solution"   
author: "Lieven Clement and Jeroen Gilis"
date: "statOmics, Ghent University (https://statomics.github.io)"  
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

# Cuckoo dataset

The common cuckoo does not build its own nest: it prefers
to lay its eggs in another birds' nest. It is known, since 1892,
that the type of cuckoo bird eggs are different between different
locations. In a study from 1940, it was shown that cuckoos return
to the same nesting area each year, and that they always pick
the same bird species to be a "foster parent" for their eggs.

Over the years, this has lead to the development of geographically
determined subspecies of cuckoos. These subspecies have evolved in
such a way that their eggs look as similar as possible as those
of their foster parents.

The cuckoo dataset contains information on 120 Cuckoo eggs,
obtained from randomly selected "foster" nests.
For these eggs, researchers have measured the `length` (in mm)
and established the `type` (species) of foster parent.
The type column is coded as follows:

- `type=1`: Meadow pipit
- `type=2`: Tree pipit
- `type=3`: Dunnock
- `type=4`: European robin
- `type=5`: White wagtail
- `type=6`: Eurasian wren

# Goal

The researchers want totest if the type of foster parent
has an effect on the average length of the cuckoo eggs.

In theory, they want to study this for all six species.
Previously, we looked at a single pairwise comparison
between the European robin and the Eurasian wren with a
t-test. Here, we will analyse all types simultaneously
with ANOVA.

Load the required libraries

```{r, message=FALSE}
library(tidyverse)
```

## Import the data

```{r, message=FALSE}
Cuckoo <- read_tsv("https://raw.githubusercontent.com/statOmics/PSLS21/data/Cuckoo.txt")
head(Cuckoo)
```

# Data tidying

It seems that the `tpye` column is a `double` rather than a `factor`. 
Let's fix this:

```{r}
Cuckoo <- Cuckoo %>%
  mutate(type = as.factor(type))
```


# Data exploration

How many birds do we have for each type?

```{r}
Cuckoo %>%
  count(type)
```

Visualize the data

```{r}
Cuckoo %>%
  ggplot(aes(x=type,y=length,fill=type)) +
    theme_bw() +
    scale_fill_brewer(palette="RdGy") +
    geom_boxplot() +
    geom_jitter(width = 0.2) +
    ggtitle("Boxplot of the length of eggs per type") +
    ylab("length (mm)") + 
    stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black")
```

# ANOVA

To study if the observed differences in average egg length 
between the different foster bird types are significant, we may perform 
an ANOVA.

## Formulate null and alternative hypothesis

The null hypothesis of ANOVA states that:
$H0$: The mean egg length is equal between the different bird types.

The alternative hypothesis of ANOVA states that:
$HA$: The mean egg length for at least one bird type is different 
from the mean egg length in at least one other bird type.

## Check assumptions

Before we may proceed with the analysis, we must make sure that all
assumptions for ANOVA are met. ANOVA has three assumptions:

1. The observations are independent of each other (in all groups)
2. The data (length) must be normally distributed (in all groups)
3. The variability within all groups is similar

The first assumption is met, as we may assume that there are no
specific patterns of correlation between the randomly selected nests.

To check the normality assumption, we will use QQ plots.

```{r}
Cuckoo %>% 
  ggplot(aes(sample=length)) +
  geom_qq() +
  geom_qq_line() + 
  facet_grid(~type)
```

There seem to be no clear deviations from normality.

The third assumption of equal variances seems to be met based on the
visualization with the boxplots (see above).

As such, we may proceed with the ANOVA analysis.

## ANOVA model

```{r}
fit <- lm(length~type, Cuckoo)
fit_anova <- anova(fit)
fit_anova
```

The p-value of the ANOVA analysis is extremely significant
(p-value = `r format(fit_anova$"Pr(>F)"[1],digits=4)`), 
so we reject the null hypothesis that the mean 
egg length is equal between the different bird types.
We can say that the mean egg length is significantly different
between at least two bird types on the 5% significance level.

Based on this analysis, we do not yet know between which particular
bird types there is a significant difference. To study this, we will
perform the Tuckey post-hoc analysis.

## Post-hoc analysis

We will perform a post-hoc analysis, to look at the difference 
in egg length between each pairwise comparison of bird types.
Importantly, with this strategy, the p-values will be correctly
adjusted for multiple testing.

The null hypothesis for each pairwise test is that there is no
difference in the mean egg length between both bird types.

The alternative hypothesis for each pairwise test states that there
is indeed a difference in the mean egg length between both bird types.

We will also calculate the confidence interval on the mean differences.

```{r,message=FALSE}
library(multcomp, quietly = TRUE)
mcp <- glht(fit, linfct = mcp(type = "Tukey"))
summary(mcp)
confint(mcp)
```

## Conclusion

We have found an extremely significant dependence 
(p-value = `r format(fit_anova$"Pr(>F)"[1],digits=4)`)
between the mean egg length and bird type on the global 
5% significance level.

The mean length of cuckoo's eggs in nests of the Eurasian
wern are smaller as compared to those from all other bird
types in the dataset:

- the meadow pipit (adjusted p-value = 0.00856, mean difference = -1.37 mm, 95% CI [-2.50323; -0.23677])
- the tree pipit (adjusted p-value = 0.00112, mean difference = -1.96 mm, 95% CI [-3.34792; -0.57208])
- the dunnock (adjusted p-value = 0.00115, mean difference = -1.99 mm, 95% CI [-3.40513; -0.57773])
- the European robin (adjusted p-value = 0.03184, mean difference = -1.45 mm, 95% CI [-2.81223; -0.07777])
- the white wagtail (adjusted p-value = 0.03184, mean difference = -1.77 mm, 95% CI [-3.16244; -0.38422])

For the other bird types, we have insufficient evidence to suggest
differences in mean length of the cuckoo bird's eggs.


