---
title: "Exercise 9.1: Non-parametric statistics on the shrimps dataset - solution"   
author: "Lieven Clement and Jeroen Gilis"
date: "statOmics, Ghent University (https://statomics.github.io)"  
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

# The shrimps dataset

Dataset on the accumulation of PCBs (Polychlorinated biphenyls)
in the adipose tissue of shrimps. PCBs are often present in coolants, and are 
know to accumulate easily in the adipose tissue of shrimps. In this experiment, 
two groups of 18 samples (each 100 grams) of shrimps each were cultivated 
in different conditions, one control condition and one condition 
where the medium was poluted with PCBs. Note that the PCB concentrations were 
measured in pg/g adipose tissue.

# Goal

The research question is; is there an effect of the 
growth condition on the PCB concentration in the adipose 
tissue of shrimps?

Load libraries:

```{r}
library(tidyverse)
```

# Import the data

```{r}
shrimps <- read.table("https://raw.githubusercontent.com/statOmics/PSLS21/data/shrimps.txt",
                      header = TRUE)
glimpse(shrimps)
```

# Data tidying

```{r}
shrimps <- shrimps %>%
  mutate(group = as.factor(group))
```

# Data exploration

The first step is to explore the data.

```{r}
shrimps %>%
  count(group)
```

Visualize the data:

```{r}
shrimps %>% 
  ggplot(aes(x=group,y=PCB.conc,fill=group)) + 
  scale_fill_brewer(palette="RdGy") +
  theme_bw() +
  geom_boxplot(outlier.shape=NA) + 
  geom_jitter(width = 0.2) +
  ggtitle("Boxplot of the PCB concentrations in two groups of shrimps") +
  ylab("PCB concentration (pg/g)") + 
  stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black")
```

We can see that for group 1 we have four very clear outliers
in the data. These values were double-checked (i.e for
typing errors), but there was no reason found to believe
that these values are incorrect.

# Analysis

As we have seen in the second tutorial, a good way for
testing the research hypothesis is to perform an unpaired
two-sample t-test to find out whether there is a significant
difference in the mean PCB concentrations between both groups
of samples. Before we can do this, we must check if all the 
required assumptions are met.

## Assumptions

1. The observations are independent of each other (in both groups)
2. The data (PCB.conc) must be normally distributed (in both groups)

The first assumption is met, as we randomly selected shrimps and
submitted them to one of two growth conditions. No underlying
correlation patterns are expected.

We can check the second assumption with a QQ-plot.

```{r}
shrimps %>% 
  ggplot(aes(sample=PCB.conc)) +
  geom_qq() +
  geom_qq_line() + 
  facet_grid(~group)
```

We clearly see that we have strong deviations from
normality. Many datapoints do not lie near the quantile-quantile
line. As such, we may conclude that our data is not normally distributed.
In addition, the visualization with the boxplots shows that the
variability is quite different between the two groups.

As such, we are not allowed to perform the t-test. 
As an alternative, we may perform a non-parametric
test, such as the Wilcoxon rank-sum test.

## Wilcoxon rank-sum test

The Wilcoxon rank-sum test (or the Mann-Whitney U test) is an important 
non-parametric data analysis method.In rank-based tests, the data 
$Y_i$ is first transformed to its ranked equivalent
\[ R_i=R(Y_i) = \#\{Y_j: Y_j\leq Y_i; j=1,\ldots, n\}. \]

Ranks are very robust to outliers. For instance, it doesn't
matter if the highest value in an hypothetical dataset has a 
value of 10 or 100; it will keep the same rank (highest rank).

Note that there might be tied values in the data, e.g., if
two samples of shrimps have an equal concentration of PCBs.
In this case, the Wilcoxon rank-sum test will compute
mid-ranks, which are calculated as follows;
\[R(Y_i) = \frac{\sum\limits_{\forall j : Y_j=Y_i}R(Y_j)}{\#{\forall j:Y_j=Y_i}}  \]
i.e., the midrank is equal to the mean of the ranks of equal
observations.

After computing the ranks and midranks, the Wilcoxon test will
compare the mean rank between both treatment groups:

```{r}
PCB_1 <- shrimps %>%
  filter(group == "1") %>%
  pull(PCB.conc)

PCB_2 <- shrimps %>%
  filter(group == "2") %>%
  pull(PCB.conc)

wilcox.test(PCB_1, PCB_2)
```

or, equivalently;

```{r}
wilcox.test(PCB.conc~group, data = shrimps)
```
 
We find that the test is significant on the 5%
significance level (p = 0.01871). The value W=88
could be calculated manually as the sum of the ranks
of the PCB concentration values that belong to 
shrimps of group 1.

### Interpretation

The interpretation of the Wilcoxon rank-sum test is
slightly more challenging than that of a t-test.
Since we are not using average values in this test,
we cannot interpret the results in terms of means
(as with the t-test). In contrast, we work with ranks,
and we will interpret the results in terms of the
*probabilistic index*.

The null hypothesis of the Wilcoxon test states that
the distributions of the probability density functions
are equal for both groups:

\[ H_0: f_1 = f_2 \]

In words: the distribution of PCB concentrations in 
shrimps are equal for both treatment conditions.

or, equivalently,

The probability that a random observation of a PCB
concentration derived from a sample of shrimps that
was grown in the control condition *is larger than or equal to*
than a random observation of a PCB concentration 
derived from a sample of shrimps that was grown in 
the other condition *is equal to 50%*.

The alternative hypothesis can be formulated as

\[ H_1: P(Y_{i1} \geq Y_{i2}) \ne 1/2 \]

In words: The probability that a random observation of a 
PCB concentration derived from a sample of shrimps that
was grown in the control condition *is larger than or equal to*
than a random observation of a PCB concentration 
derived from a sample of shrimps that was grown in 
the other condition *is not equal to 50%*.

We can manually calculate this probability, that is 50% 
under the null hypothesis, based on the observed test statistic.

```{r}
n1 <- n2 <- 18 #18 observations in each group
WObs <- wilcox.test(PCB.conc~group, data=shrimps)$statistic # get the observed test statistic
WObs/(n1*n2)
```

We can see that the point estimation of this probability 
is `r round(WObs/(n1*n2),3)*100`%.

We can interpret this as follows;

The probability that the PCB concentration in a random
shrimp that was grown in the control condition is 
greater than or equal to the PCB concentration in a random
shrimp that was grown in the treatment condition is
equal to 27.2%.

This chance is significantly different from 50% on the
5% significance level (p = 0.01871).






