---
title: "Exercise 5.1: Hypothesis testing on the captopril dataset - solution"   
author: "Lieven Clement and Jeroen Gilis"
date: "statOmics, Ghent University (https://statomics.github.io)"  
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

In this exercise, we will see some basics in hypothesis testing and more 
specifically on t-tests. As an example, we will work with the
captopril dataset that we already explored in the exercise on data exploration.

The goal is to answer these two research questions;

1. Is the average systolic blood pressure
before captopril treatment (SBPb) higher than 149 mmHg?

2. Is the average SBP before captopril treatment 
significantly different from the average SBP after
captopril treatment?

First, load the required R libraries:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Import the data

```{r}
(captopril <- read.table("https://raw.githubusercontent.com/statOmics/PSLS21/data/captopril.txt", 
                         header = TRUE, 
                         sep = ","))
```

# Data exploration

Before we start delving into the data in order to solve 
our research hypothese, it is always a good idea to first
have a look at the data. Our dataset looks like this;

```{r}
head(captopril)
```

We have 15 patients, for which we have measure the systolic 
blood pressure and diastolyic blood pressure, before and after
treatment with the captopril drug.

Note that the dataset is not in the tidy format. We could tidy the data using
the following code.

```{r}
captopril %>% 
    gather(type,bp,-id) %>%
    filter(type%in%c("SBPa","SBPb")) %>%
    mutate(id = as.factor(id))
    # we are only interested in the systolic blood pressure
```

We can visualize the entire dataframe in an informative way 
with boxplots;

```{r}
captopril %>% 
  gather(type,bp,-id) %>%
  filter(type%in%c("SBPa","SBPb")) %>%
  ggplot(aes(x=type, y=bp, fill=type)) + 
      scale_fill_brewer(palette="RdGy") +
      theme_bw() +
      geom_boxplot(outlier.shape=NA) + 
      geom_jitter(width = 0.2) +
      ggtitle("Boxplot of blood pressure measures before and after treatment") +
      ylab("blood pressure (mmHg)") +
      stat_summary(fun=mean, geom="point", shape=5, size=3, color="black", fill="black")
```

Clearly, it seems that on average the measurements
after treatment are lower than those before treatment.
But is this difference **significant**? To answer this
question, we will need to perform hypothesis tests.
Let's start of with question 1.

# Question 1

Is the average systolic blood pressure (SBP)
before captopril treatment higher than 149 mmHg?

In the data exploration of the NHANES dataset we have set up a reference
interval, i.e. an interval that is expected to hold 95% of
the SBP values of healthy individuals. We found the interval
of [93;149] mmHg. (If you did not get to this point of the script
yet, don't worry!).

To test the effect of the captopril on subjects with hypertension
(patients), we need to find a group of patients that have
elevated SBP levels, higher than 149 mmHg.Therefore, we want
to test whether the patients in the captopril study indeed have on average a SBP level
that is greater than 149 mmHg. 
We can assess this research hypothesis using a one sample t-test.

## Assess the assumptions

Before we can perform a t-test, we must check that the required
assumptions are met!

1. The observations are independent of each other
2. The data (SBPb) must be normally distributed

For the first assumption requires us to think about the data. 
Are there any underlying correlation structures (that we know of)
in the data? For instance, if all the 15 subjects are members of
the same family, we expect that the data will give us a good 
representation of the underlying population of interest, i.e., 
all past, present and future patients with elevated SBP levels.

In this dataset, we have no reason to believe that this 
assumption was violated; we may we have assume 15 unrelated, 
"random"  patients with elevated SBP levels.

We can assess the second assumption with a quantile-quantile plot.

```{r}
captopril %>%
  ggplot(aes(sample=SBPb)) +
  geom_qq() +
  geom_qq_line()
```

We can see that all of the data lies nicely around the quantile-quantile
line (black line). As such, we may conclude that our data is normally 
distributed.

## Hypothesis test

Here, we will test if mean SBPb is significantly higher than 149 mmHg.More 
specifically, we will test the null hypothesis;

$H0:$ the mean SBPb is equal to 149 mmHg

versus the alternative hypothesis;

$HA:$ the mean SBPb is greater than 149 mmHg

```{r}
output1 <- t.test(captopril$SBPb, 
                  mu=149,
                  alternative = "greater",
                  conf.level = 0.95)
output1
```

## Conclusion

When writing a conclusion on your research hypothesis,
it is very important to be precise, concise, and complete.

An example of such a conclusion for our research question
is given below:

The mean SBP of patients before treatment with captopril
is significantly higher (p=`r round(unname(output1$p.value),5)`) than the 
upper bound of the reference interval (147 mmHg) on the 5% signifcance level. 
The mean SBPb equals `r round(unname(output1$estimate),2)` mmHg with 
a 95% confidence interval of [`r round(output1$conf.int[c(1,2)],2)`]).

As we have seen in the theory class, the 95% confidence 
interval can be interpreted as;

With 95% confidence we can state that the interval 
[`r round(output1$conf.int[c(1,2)],2)`]
contains the true average of SBP of diseased patient before
treatment with captopril.

# Question 2

Is the average SBP before captopril treatment different from the average 
SBP after captopril treatment?

As the data is paired, there will be a strong correlation between the BP values 
before and after treatment of each individual patient. We can show this
with a scatterplot.

```{r}
captopril %>% 
  ggplot(aes(x=SBPb,y=SBPa)) +
    geom_point() +
    ggtitle("correlation between SBPb and SBPa") +
    ylab("SBPa (mmHg)") +
    xlab("SBPb (mmHg)")
```

We clearly see that if a patient's SBPb value is high, its
SBPa value will be comparatively high as well.

## Check the assumptions

The paired t-test has 2 assumptions:

1. The observations are independent of each other (in both groups)

2. The data (SBPb and SBPa) must be normally distributed (in both groups)

Additionally, we must check if the variances are similar for both groups.
If so, we can use a t-test with a pooled variance (see theory).
If not, we must rely on the Welch t-test, which can deal with
unequal variances.

The first assumption is met given the experimental design.

Secondly, we assess if the data are normally distributed.

```{r}
captopril %>%
  ggplot(aes(sample=SBPa)) +
  geom_qq() +
  geom_qq_line()

captopril %>%
  ggplot(aes(sample=SBPb)) +
  geom_qq() +
  geom_qq_line()
```

We can see that all of the data lies nicely around the quantile-quantile
line. As such, we may conclude that our data is normally distributed.

For the third assumption, we must compare the within-group variability of both 
groups. We can do this visually with boxplots.

```{r}
captopril %>% 
  select(SBPb,SBPa) %>%
  gather(type,bp) %>% 
  ggplot(aes(x=type,y=bp,fill=type)) + 
    scale_fill_brewer(palette="RdGy") +
    theme_bw() +
    geom_boxplot(outlier.shape=NA) + 
    geom_jitter(width = 0.2) +
    ggtitle("Boxplot of different blood pressure measures") +
    ylab("blood pressure (mmHg)") + stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black")
```

As a measure of variability, we may take the height
of each boxplot's box. This is the interval between
the 25% and 75% quantile. Here we can see that this
interval, as well as the length of the whiskers, is
approximately equal for both groups. When the sample
sizes are small (as is the case here, we speak about deviation
from equality if one height is more than 2 or 3 times
larger/smaller than that of the other group. 

As all three assumptions are met we may continue with
performing the unpaired two-sample t-test.

## Hypothesis test

As such, we will now perform a `paired` t-test.

```{r}
output2 <- t.test(captopril$SBPb, captopril$SBPa, paired = TRUE)
output2
```

Clearly, by correctly stating that the data is paired,
we have gained a lot of statistical power for rejecting
the null hypothesis that the true differenc in means is
equal to 0. The p-value (p = `r round(unname(output2$p.value),5)`) has now 
become extremely significant. Note that the 95% CI has become narrower!

## Conclusion

We may conclude that, on the 5% significance level, the mean 
SBP levels of patients before captopril treatment is 
extremely significantly (p = `r round(unname(output2$p.value),5)`) higher than the mean 
SBP levels of patients after captopril treatment. The SBP levels 
are on average `r round(unname(output2$estimate),2)` mmHg
higher before treatment than after treatment (95% CI [`r round(output2$conf.int[c(1,2)],2)`]).

# Alternative solution: One-sample t-test on the difference

On final thing; performing a paired two-sample t-test is
analogous to performing a one-sample t-test on the difference
between both groups.

This can be easily seen from the output of the paired two-sample
t-test. The alternative hypothesis $HA$ there states that 
the "true difference in means is not equal to 0". So internally,
R will actually perform a one-sample t-test on the difference, and
check whether or not the true mean difference is equal to 0.
We can also set this up manually.

```{r}
bp_diff <- captopril %>%
  mutate(bp_diff = SBPb-SBPa) %>%
  select(bp_diff)

t.test(bp_diff,mu=0)
```

Indeed, the output is equivalent to that of the paired two-sample t-test.







